### 2015-03-17 doc/thesis/system.tex
Minor improvements to Chapter 4

In this chapter we present the formalization of Typed Lua's type system.
Besides its practical contributions, Typed Lua also has some interesting
contributions to the field of optional type systems for scripting
languages.
They are novel type system features that let Typed Lua cover several Lua idioms
and features, such as refinement of tables, multiple assignment, multiple return values,
and optional parameters.

@@

Union types $T_{1} \cup T_{2}$ represent types that can hold a value
of two different types.

@@

Typed Lua includes filter types as a way to discriminate the type of local
variables inside conditions.
In Section \ref{sec:fap} we show in more detail how our type system
uses them to formalize the \texttt{type} predicates that we mentioned
in Section \ref{sec:unions}.

Typed Lua includes projection types as a way to project
unions of tuple types into unions of first-level types.
In Section \ref{sec:fap} we show in more detail how our type system
uses them as a mechanism for handling unions of tuple types,
when they appear on the right-hand side of the declaration of local variables,
as we mentioned in Section \ref{sec:unions}.
We also show how this feature allows our type system to constrain
the type of a local variable that depends on the type of another local variable.

@@

Simply promoting the type of each table value to its supertype would
not overcome this limitation, as it still would give to the table constructor
a regular table type without covariant mutable fields.
Thus, programmers would not be able to use the table constructor to
initialize a variable with a table type that includes an optional field.
Using the previous example,
\begin{align*}
& \{``x":\Integer, ``y":\Integer\}_{closed} \not\subtype \\
& \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
\end{align*}

@@

In the next section we will show in more detail how our type system
uses these tags to handle the refinement of table types and idiomatic
object-oriented code.

@@

Figure \ref{fig:syntax} presents the abstract syntax of core Typed Lua.
It splits the syntactic categories as follows:
$s$ are statements, $e$ are expressions, $l$ are left-hand values,
$k$ are literal constants, $el$ are expression lists,
$me$ are expressions with multiple results, $a$ are function and method applications,
$f$ are function declarations, $pl$ are parameter lists,
$id$ are variable names, $T$ are first-level types, and $S$ are second-level types.
The notation $\overline{id{:}T}$ denotes the list $id_{1}{:}T_{1}, ..., id_{n}{:}T_{n}$,
where $n$ can be zero.

@@

As an example, lets assume that $x$ and $y$ are variables in the
environment with types $\Integer$ and $\String$.
Let us see how our type system type checks the following assignment
\[
x, y = 1, ``foo"
\]

First, our type system type checks the expression list in the right-hand
side of the assignment.
In our example, the right-hand side of the assignment has type
$1 \times ``foo" \times \Nil{*}$.
Note that our type system includes the type $\Nil{*}$ to replace
missing values.
The rules that type check expression lists introduce the type
$\Nil{*}$ to let the right-hand side produce fewer values than
expected in the left-hand side.
Our example uses the rule \textsc{T-EXPLIST1} which is the rule
that type checks an expression list where all expressions can
only produce a single value:
\[
\begin{array}{c}
\mylabel{T-EXPLIST1}\\
\dfrac{\env_{i}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
       n = |\;\overline{e}\;|}
      {\env_{1}, \penv \vdash \overline{e}:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{m}}
\end{array}
\]

@@

Then, our type system type checks the list of left-hand values.
In our example, the left-hand side of the assignment has type 
$\Integer \times \String \times \Value{*}$.
Note that our type system uses the type $\Value{*}$ to discard extra values.
The rule that type checks lists of left-hand values introduces the type
$\Value{*}$ to let the right-hand side produce more values than
expected in the left-hand side.
Our example uses the rule \textsc{T-LHSLIST} which is the rule that type
checks a list of left-hand values:
\[
\begin{array}{c}
\mylabel{T-LHSLIST}\\
\dfrac{\env_{i}, \penv \vdash l_{i}:T_{i}, \env_{i+1} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
       n = |\;\overline{l}\;|}
      {\env_{1}, \penv \vdash \overline{l}:T_{1} \times ... \times T_{n} \times \Value{*}, \env_{m}}
\end{array}
\]

@@

We also use the rule \textsc{T-APPLY1} as the base case for the rules
that handle the cases where function applications are either statements
or expressions that produce only one value.
The rule \textsc{T-STMAPPLY1} discards the produced values,
while the rule \textsc{T-EXPAPPLY1} uses the auxiliary function
\emph{first} to ensure that only one value is produced:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-STMAPPLY1}\\
\dfrac{\env_{1}, \penv \vdash e(el):S, \env_{2}}
      {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{0},\env_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{T-EXPAPPLY1}\\
\dfrac{\env_{1}, \penv \vdash e(el):S, \env_{2}}
      {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{1}:first(S), \env_{2}}
\end{array}
\end{array}
\]

@@

Our type system also catches arity mismatch.
For instance, let us assume that $f$ has type
$\String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Void \rightarrow \Integer{*}$.
The function call
\[
f(``foo")
\]
type checks through the rule \textsc{T-APPLY1}, because
\[
``foo" \times \Nil{*} \lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Void
\]
but the function call
\[
f(``foo",1,2,3)
\]
does not type check through the rule \textsc{T-APPLY1}, because
\[
``foo" \times 1 \times 2 \times 3 \times \Nil{*} \not\lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Void
\]

@@

Aliasing an \emph{unique} or an \emph{open} table type can produce
either a \emph{closed} or a \emph{fixed} table type, depending on
the context that we are using a variable.
As we mentioned in Sections \ref{sec:oop} and \ref{sec:subtyping},
we need \emph{fixed} table types to type classes in object-oriented programming.
In the implementation we fix the aliasing of \emph{unique} and \emph{open}
table types that appear in a return statement, and in other cases we
close the aliasing of \emph{unique} and \emph{open} table types.
However, in the formalization we chose to define this behavior not
deterministically, as it makes easier the presentation of this behavior.

@@

As an example, let us assume that \emph{idiv} and \emph{print} are functions
in the environment.
The function \emph{idiv} has type
\[
\Integer \times \Integer \times \Void \rightarrow (\Integer \times \Integer \times \Void) \sqcup (\Nil \times \String \times \Void)
\]
As we mentioned in Section \ref{sec:unions}, \emph{idiv} performs
integer division.
In case of success, it returns two integers: the result and the remainder.
In case of failure, it returns $\Nil$ plus an error message that indicates
the error.
The function \emph{print} is a variadic function of type
$\Value{*} \rightarrow \Void$.
Let us see how our type system type checks the following program:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; q, r = idiv(1, 2) \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$\mathbf{if} \; q \; \mathbf{then} \; \lfloor print(q + r) \rfloor_{0} \; \mathbf{else} \; \lfloor print(``ERROR: " \; .. \; r) \rfloor_{0}$}
\end{tabular}
\end{center}

@@

In the rule that type checks the declaration of unannotated variables,
our type system uses the pair $(x,S_{2})$ to bound a union of
second-level types $S_{2}$ to a variable $x$ in the projection
environment $\penv$.
In our example, declaring $q$ and $r$
bounds the projection type $\pi_{1}^{x}$ to $q$ and
bounds the projection type $\pi_{2}^{x}$ to $r$,
where the projection variable $x$ bounds to 
\[
(\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*})
\]
in the projection environment $\penv$.
Rule \textsc{T-LOCAL2} captures this intuition:
\[
\begin{array}{c}
\mylabel{T-LOCAL2}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash el:S_{1}, \env_{2}, (x,S_{2})\\
       \env_{2}[id_{1} \mapsto infer(S_{1},1), ..., id_{n} \mapsto infer(S_{1},n)], \penv[x \mapsto S_{2}] \vdash s, \env_{3} \;\;\;
       n = |\;\overline{id}\;|  
       \end{array}}
      {\env_{1}, \penv \vdash \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s, \env_{3} - \{\overline{id}\}}
\end{array}
\]

@@

After assigning projection types to $q$ and $r$, reading $q$ will
use the projection type $\pi_{1}^{x}$ to project the type of $q$
into the union type $\Integer \cup \Nil$, while reading $r$ will
use the projection type $\pi_{2}^{x}$ to project the type of $r$
into the union type $\Integer \cup \String$.
However, we want to discriminate these variables to check whether
the function call returned with success.

>>>>

In this chapter we present the abstract syntax of Typed Lua types,
the subtyping rules, and the most interesting typing rules.
Besides its practical contributions, Typed Lua also has some interesting
contributions to the field of optional type systems for scripting
languages.
They are novel type system features that let Typed Lua cover several Lua idioms
and features, such as refinement of tables, multiple assignment, and multiple return values.

@@

Union types $T_{1} \cup T_{2}$ represent data types that can hold a value
of two different types.

@@

Typed Lua includes filter types as a way to discriminate the type of local
variables inside conditions.
Our type system uses filter types to formalize the \texttt{type} predicates
that we mentioned in Section \ref{sec:unions}.
This means that \texttt{type} predicates use filter types of the form
$\phi(T_{1},T_{2})$ to discriminate local variables that are bound to
union types.
In a filter type $\phi(T_{1},T_{2})$, $T_{1}$ is the original type and
$T_{2}$ is the discriminated type.

Typed Lua includes projection types as a way to project
unions of tuple types into unions of first-level types.
In Section \ref{sec:fap} we will show in more detail how our type system
uses them as a mechanism for handling unions of tuple types,
when they appear in the right-hand side of the declaration of local variables,
as we mentioned in Section \ref{sec:unions}.
We also show how this feature allows our type system to constrain
the type of a local variable that depends on the type of another local variable.

@@

Simply promoting the type of each table value to its supertype would
not overcome this limitation, as it still would give to the table constructor
a \emph{closed} table type without covariant mutable fields.
Thus, programmers would not be able to use the table constructor to
initialize a variable with a table type that includes an optional field.
Using the previous example,
\begin{align*}
& \{``x":\Integer, ``y":\Integer\}_{closed} \not\subtype \\
& \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
\end{align*}

@@

In the next section we will show in more detail how our type system
uses these tags to handle the refinement of table types.

@@

Figure \ref{fig:syntax} presents the abstract syntax of core Typed Lua.
It splits the syntactic categories as follows:
$s$ are statements, $e$ are expressions, $l$ are left-hand values,
$k$ are literal constants, $el$ are expression lists,
$me$ are expressions with multiple results, $a$ are function and method applications,
$f$ are function declarations, $pl$ are parameter lists,
$id$ are variable names, $T$ are first-level types, and $S$ are second-level types.
The notation $\overline{id{:}T}$ denotes the list $id_{1}{:}T_{1}, ..., id_{n}{:}T_{n}$.

@@

As an example, lets assume that $x$ and $y$ are variables in the
environment with types $\Integer$ and $\String$.
Let us see how our type system type checks the following assignment:
\[
x, y = 1, ``foo"
\]

First, our type system type checks the expression list in the right-hand
side of the assignment.
In our example, the right-hand side of the assignment has type
$1 \times ``foo" \times \Nil{*}$.
Note that our type system includes the type $\Nil{*}$ to replace
missing values.
The rules that type check expression lists introduce the type
$\Nil{*}$ to let the right-hand side produce fewer values than
expected in the left-hand side.
Our example uses the rule \textsc{T-EXPLIST1} to type check
the right-hand side of the assignment.
The rule \textsc{T-EXPLIST1} is the rule that type checks an
expression list where all expressions can only produce a single value:
\[
\begin{array}{c}
\mylabel{T-EXPLIST1}\\
\dfrac{\env_{1}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
       n = |\;\overline{e}\;|}
      {\env_{1}, \penv \vdash \overline{e}:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{m}}
\end{array}
\]

@@

After type checking the right-hand side, our type system type checks the list
of left-hand values.
In our example, the left-hand side of the assignment has type 
$\Integer \times \String \times \Value{*}$.
Note that our type system uses the type $\Value{*}$ to discard extra values.
The rule that type checks lists of left-hand values introduces the type
$\Value{*}$ to let the right-hand side produce more values than
expected in the left-hand side.
Our example uses the rule \textsc{T-LHSLIST} to type
checks a list of left-hand values:
\[
\begin{array}{c}
\mylabel{T-LHSLIST}\\
\dfrac{\env_{1}, \penv \vdash l_{i}:T_{i}, \env_{i+1} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n+1}) \;\;\;
       n = |\;\overline{l}\;|}
      {\env_{1}, \penv \vdash \overline{l}:T_{1} \times ... \times T_{n} \times \Value{*}, \env_{m}}
\end{array}
\]

@@

We also use the rule \textsc{T-APPLY1} as the base case for the rules
that handle the cases where function applications are either statements
that produce no value or expressions that produce only one value.
The rule \textsc{T-STMAPPLY1} discards the produced values,
while the rule \textsc{T-EXPAPPLY1} uses the auxiliary function
\emph{first} to ensure that only one value is produced:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-STMAPPLY1}\\
\dfrac{\env_{1}, \penv \vdash e(el):S, \env_{2}}
      {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{0},\env_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{T-EXPAPPLY1}\\
\dfrac{\env_{1}, \penv \vdash e(el):S, \env_{2}}
      {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{1}:first(S), \env_{2}}
\end{array}
\end{array}
\]

@@

Our type system also catches arity mismatch.
To do that, we end the input type of a funcion with type $\Void$
instead of $\Value{*}$.
For instance, let us assume that $f$ has type
$\String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Void \rightarrow \Integer{*}$.
The function call
\[
f(``foo")
\]
type checks through the rule \textsc{T-APPLY1}, because
\[
``foo" \times \Nil{*} \lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Void
\]
but the function call
\[
f(``foo",1,2,3)
\]
does not type check through the rule \textsc{T-APPLY1}, because
\[
``foo" \times 1 \times 2 \times 3 \times \Nil{*} \not\lesssim \String \times (\Integer \cup \Nil) \times (\Integer \cup \Nil) \times \Void
\]

@@

Aliasing an \emph{unique} or an \emph{open} table type can produce
either a \emph{closed} or a \emph{fixed} table type, depending on
the context that we are using a variable.
As we mentioned in Sections \ref{sec:oop} and \ref{sec:subtyping},
we need \emph{fixed} table types to type classes in object-oriented programming.
In the implementation we fix the aliasing of \emph{unique} and \emph{open}
table types that appear in a top-level return statement, and in other cases we
close the aliasing of \emph{unique} and \emph{open} table types.
However, in the formalization we chose to define this behavior not
deterministically, as it makes easier the presentation of this behavior.

@@

As an example, let us assume that \emph{idiv} and \emph{print} are functions
in the environment.
The function \emph{idiv} has type
\[
\Integer \times \Integer \times \Void \rightarrow (\Integer \times \Integer \times \Void) \sqcup (\Nil \times \String \times \Void)
\]
As we mentioned in Section \ref{sec:unions}, \emph{idiv} performs
integer division.
In case of success, it returns two integers: the result and the remainder.
In case of failure, it returns $\Nil$ plus an error message that describes
the error.
The function \emph{print} is a variadic function of type
$\Value{*} \rightarrow \Void$.
Let us see how our type system type checks the following program:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; q, r = idiv(1, 2) \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$\mathbf{if} \; q \; \mathbf{then} \; \lfloor print(q + r) \rfloor_{0} \; \mathbf{else} \; \lfloor print(``ERROR: " \; .. \; r) \rfloor_{0}$}
\end{tabular}
\end{center}

@@

In the rule that type checks the declaration of unannotated variables,
our type system uses the pair $(x,S_{2})$ to bound a union of
second-level types $S_{2}$ to a variable $x$ in the projection
environment $\penv$.
In our example, declaring $q$ and $r$
bounds the projection type $\pi_{1}^{x}$ to $q$ and
bounds the projection type $\pi_{2}^{x}$ to $r$,
where the projection variable $x$ bounds to 
\[
(\Integer \times \Integer \times \Nil{*}) \sqcup (\Nil \times \String \times \Nil{*})
\]
in the projection environment $\penv$.
Rule \textsc{T-LOCAL2} illustrates this intuition:
\[
\begin{array}{c}
\mylabel{T-LOCAL2}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash el:S_{1}, \env_{2}, (x,S_{2})\\
       \env_{2}[id_{1} \mapsto infer(S_{1},1), ..., id_{n} \mapsto infer(S_{1},n)], \penv[x \mapsto S_{2}] \vdash s, \env_{3} \;\;\;
       n = |\;\overline{id}\;|  
       \end{array}}
      {\env_{1}, \penv \vdash \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s, \env_{3} - \{\overline{id}\}}
\end{array}
\]

@@

After assigning projection types to $q$ and $r$, reading $q$ will
use the projection type $\pi_{1}^{x}$ to project the type of $q$
into the union type $\Integer \cup \Nil$, while reading $r$ will
use the projection type $\pi_{2}^{x}$ to project the type of $r$
into the union type $\Integer \cup \String$.
Now, we may want to discriminate these variables to check whether
the function call returned with success.
